---
title: "Metiers and things"
author: Einar Hjörleifsson
description: "In my ignorance I assumed that the metiers, targets and gears in the ICES vocabulary were hierarchical. In the sense that if one has a **valid** metier level 6, one could derive **valid** metier level 5, target and gear from that. Here a little script is provided that shows that this is not quite the case."
date: "2025-07-17"
categories: [code, rtip]
echo: true
---

## First some convenience wrappers

Not really needed here, but I use these functions elsewhere:

```{r}
library(tidyverse)
#' Derive gear from metier 5 or 6
#' 
#' Obtain gear type (sometimes wrongly called metier 4?) from metier5 or  6
#'
#' @param x A character vector
#'
#' @return A vector
#' @export
#'
rb_gear_from_metier <- function(x) {
  x |> stringr::str_split("_") |> purrr::map_chr(1)
}
#' Derive target from metier 5 or 6
#'
#' Not to be confused with metier 5
#' 
#' @param x A character vector
#'
#' @return A vector
#' @export
#'
rb_target_from_metier <- function(x) {
  x |> stringr::str_split("_") |> purrr::map_chr(2)
}
#' Derive metier 5 from metier 6
#'
#' Not to be confused with target assemblage
#' 
#' @param x A character vector
#'
#' @return A vector
#' @export
#'
rb_met5_from6 <- function(x) {
  sapply(strsplit(x, "_"), function(x) paste(x[1:2], collapse = "_"))
}
```

## The test

Steps:

* Get all the metier level 6 from ICES
* Derive metier level5, target and gear from metier level 6
* Test if derived values are available (valid) using ICES vocabulary

```{r}
m6 <- 
  icesVocab::getCodeList("Metier6_FishingActivity") |> 
  as_tibble() |> 
  select(key = Key, desc = Description, dep = Deprecated) |> 
  # get rid of some mess
  mutate(desc = stringr::str_remove(desc, ", see.*device")) |> 
  mutate(gear = rb_gear_from_metier(key),
         target = rb_target_from_metier(key),
         met5 = rb_met5_from6(key)) |>  
  mutate(
    v_gear = 
      case_when(gear %in% icesVocab::getCodeList("GearType")$Key ~ TRUE,
                .default = FALSE),
    v_target = 
      case_when(target %in% icesVocab::getCodeList("TargetAssemblage")$Key ~ TRUE,
                .default = FALSE),
    v_met5 = 
      case_when(met5 %in% icesVocab::getCodeList("Metier5_FishingActivity")$Key ~ TRUE,
                .default = FALSE)
  )
m6 |> count(v_gear, v_target, v_met5)
m6 |> 
  filter(!v_gear | !v_target | !v_met5) |> 
  select(key, dep, ends_with("gear"), ends_with("target"), ends_with("met5")) |> 
  knitr::kable(caption = "Something wrotten in the State of Denmark")
```


## Check going from metier level 5

Steps:

* Get all the metier level 5 from ICES
* Derive target and gear from metier level 5
* Test if derived values are available (valid) using ICES vocabulary

```{r}
m5 <- 
  icesVocab::getCodeList("Metier5_FishingActivity") |> 
  as_tibble() |> 
  select(key = Key, desc = Description, dep = Deprecated) |> 
  mutate(gear = rb_gear_from_metier(key),
         target = rb_target_from_metier(key)) |>  
  mutate(v_gear = case_when(gear %in% icesVocab::getCodeList("GearType")$Key ~ TRUE,
                            .default = FALSE),
         v_target = case_when(target %in% icesVocab::getCodeList("TargetAssemblage")$Key ~ TRUE,
                              .default = FALSE))
m5 |> count(v_gear, v_target)
m5 |> 
  filter(!v_gear | !v_target) |> 
  select(key, dep, ends_with("gear"), ends_with("target"), ends_with("met5")) |> 
  knitr::kable(caption = "Something wrotten in the State of Denmark")
```

## The Icelandic gear table

```{r}
gear_mapping <- nanoparquet::read_parquet(here::here("data/gear/gear_mapping.parquet"))
gear_mapping |> knitr::kable(caption = "Icelandic gear mapping")
```

So which gears do not pass the test:

```{r}
gear_mapping |> 
  filter(!is.na(gear)) |> 
  mutate(
    v_gear = 
      case_when(gear %in% icesVocab::getCodeList("GearType")$Key ~ TRUE,
                .default = FALSE),
    v_target = 
      case_when(target %in% icesVocab::getCodeList("TargetAssemblage")$Key ~ TRUE,
                .default = FALSE),
    v_met5 = 
      case_when(met5 %in% icesVocab::getCodeList("Metier5_FishingActivity")$Key ~ TRUE,
                .default = FALSE),
    v_met6 = 
      case_when(met6 %in% icesVocab::getCodeList("Metier6_FishingActivity")$Key ~ TRUE,
                .default = FALSE)
  ) |> 
  filter(!v_gear | !v_target | !v_met5 | !v_met6) |> 
  knitr::kable(caption = "Icelandic gears that pass metier level 5 and 6, but not gear and target")
```

So here we have a bit of a catch22, metier 5 and 6 valid but not gear (except DRB) and target.
