---
title: "Observer trips"
---

## Data source

Two data-sources:

* Directly from Fiskistofa
  * Contains observer trips that have been matched to landings id
  * Time range: 2016-01-01 to 2019-12-31
  * Number of observations (trips): 1128
* From Fiskistofa via Daði:
  * Contains observer trips that have not been matched to landings id
  * Time range: 2017-01-02 to 2022-10-24
  * Number of observations (trips): 1371


### Directly from Fiskistofa

```{r}
library(data.table)
library(tidyverse)
library(omar)
con <- connect_mar()
lh_date <- function(x) {
  if_else(str_detect(x, "\\."),
          dmy(x),
          janitor::excel_numeric_to_date(as.numeric(x)))
}
gear_code <- 
  tbl_mar(con, "agf.aflagrunnur_v") |> 
  select(gid = veidarfaeri,
         gear = veidarfaeri_heiti) |> 
  distinct() |> 
  collect() |> 
  arrange(gid)

obs1 <-
  readxl::read_excel(here::here("data-external/SJOFERDIR_01012016_31122019.xlsx"), col_types = "text") |>
  rename(vid = 1,
         gear = 2,
         t1 = 3,
         t2 = 4,
         .id = 5,
         datel = 6) |> 
  mutate(vid = as.integer(vid),
         t1 = lh_date(t1),
         t2 = lh_date(t2),
         .id = as.integer(.id),
         datel = lh_date(datel)) |> 
  mutate(t1 = if_else(t1 == ymd("2011-11-25"),
                      ymd("2017-11-25"),
                      t1)) |> 
  arrange(vid, t1, t2) |> 
  mutate(.rid = 1:n(),
         .cid = ramb::rb_event(paste(vid, t1, t2))) |> 
  select(.cid, everything())
```

### Via Daði

```{r}
obs2 <-
  read_csv(here::here("data-external/SJOFERDIR_01012017_25102022.csv")) |> 
  rename(vid = 1,
         t1 = 2,
         t2 = 3) |> 
  mutate(vid = as.integer(vid),
         t1 = dmy(t1),
         t2 = dmy(t2)) |> 
  filter(vid > 0) |> 
  arrange(vid, t1, t2) |> 
  mutate(.rid = 1:n(),
         .cid = ramb::rb_event(paste(vid, t1, t2))) |> 
  select(.cid, everything())
```

## Can we automatically obtain landings id?

In dataset 1 we have reported t2 and the landings date, the latter presumably from the landings database. Let's check the match/difference in the two:

```{r}
obs1 |> 
  select(vid, t1, t2, datel) |> 
  distinct() |> 
  mutate(dt = difftime(datel, t2, units = "days")) |> 
  count(dt) |> 
  mutate(p = round(n / sum(n), 3)) |> 
  knitr::kable(caption = "Difference in reported trip end date and landings date")
```

So in 86% of cases the two dates match, in another 10% the date landings recorded in the landings database is 1 day after the reported drip end date.

### Trial via nearest date match

```{r}
match_nearest_date <- function(lb, ln) {
  
  lb.dt <-
    lb %>%
    select(vid, datel) %>%
    distinct() %>%
    setDT()
  
  ln.dt <-
    ln %>%
    select(vid, datel) %>%
    distinct() %>%
    mutate(dummy = datel) %>%
    setDT()
  
  res <-
    lb.dt[, date.ln := ln.dt[lb.dt, dummy, on = c("vid", "datel"), roll = "nearest"]] %>%
    as_tibble()
  
  lb %>%
    left_join(res,
              by = c("vid", "datel")) %>%
    left_join(ln %>% select(vid, date.ln = datel, gid_ln = gid, .id),
              by = c("vid", "date.ln"))
  
}
# try first the obs1 data, prentend they are logbook data
lb <- 
  obs1 |> 
  # use t2 as proxy for date landed (are pretending" we do not have it)
  select(.cid, .rid, vid, t1, datel = t2, .id_org = .id, gear)
ln <- 
  tbl_mar(con, "agf.aflagrunnur") |>
  filter(between(year(londun_hefst), 2016, 2019),
         magn > 0) |> 
  select(vid = skip_numer, 
         datel = londun_hefst, 
         gid = veidarfaeri,
         .id = londun_id) |> 
  collect() |> 
  mutate(datel = as_date(datel)) |>
  distinct() |> 
  arrange(vid, datel) |> 
  filter(vid > 0)
m <- 
  match_nearest_date(lb, ln)
```

First thing to note is that we have more number of records than the original (obs `r nrow(obs1)` vs m `r nrow(m)`). Let's check why:

```{r}
m |> 
  group_by(.rid) |> 
  mutate(n = n()) |> 
  filter(n > 1) |> 
  knitr::kable(caption = "Records with row expansions")
```

In most cases this seems to be just an issue with the many-to-many relationship. We could try summarise the data such that only one landing id (the minumum used here) is used for what constitutes a trip:

```{r}
obs1_min <- 
  obs1 |> 
  group_by(.cid) |> 
  mutate(.id_min = min(.id)) |> 
  ungroup() |> 
  select(.cid, vid, t1, t2, datel, .id_min) |> 
  distinct()
ln_min <- 
  ln |> 
  group_by(vid, datel) |> 
  mutate(.id_min = min(.id)) |> 
  ungroup() |> 
  select(-.id) |> 
  distinct()
lb <- 
  obs1_min |> 
  # use t2 as proxy for date landed (are pretending" we do not have it)
  rename(.id_org = .id_min)
m <- 
  match_nearest_date(lb, ln_min |> rename(.id = .id_min))
m |> 
  filter(.id_org != .id) |> 
  knitr::kable(caption = "Records with mismatching landings id depending on approach used")
ID <- 
  m |> 
  filter(.id_org != .id) |> 
  pull(.id_org)
ln_dt <- # double test
  tbl_mar(con, "agf.aflagrunnur") |> 
  filter(londun_id %in% ID) |> 
  select(.id = londun_id,
         vid = skip_numer,
         datel = londun_hefst,
         fisktegund,
         magn,
         magn_oslaegt) |> 
  collect()
m |> 
  filter(.id_org != .id) |> 
  left_join(ln_dt)
```

Not quite clear what is going on.

```{r}
m |> 
  mutate(what = case_when(is.na(.id) ~ "NA",
                          .id == .id_org ~ "same",
                          .id != .id_org ~ "diff",
                          .default = "other")) |> 
  mutate(dt = difftime(datel, date.ln, units = "days")) |> 
  filter(what == "diff")
```

```{r}
# first line
ID <- c(499646, 499470)
m |> 
  filter(.id_org %in% ID | .id %in% ID)
ln |> 
  filter(.id %in% ID) |> 
  glimpse()
# second line
ID <- c(500124, 500073)
m |> 
  filter(.id_org %in% ID | .id %in% ID)
ln |> 
  filter(.id %in% ID) |> 
  glimpse()
m |> 
  arrange(vid, date.ln, .id) |> 
  left_join(ln |> 
              select(vid, date.ln = datel, .id_auto = .id) |> 
              arrange(vid, date.ln, .id_auto)) |> 
  filter(.id != .id_auto)
# think may be better to use bind_rows and then find a match
m
```

## Number of species by trip type

```{r}
ln <-
  tbl_mar(con, "agf.aflagrunnur") |>
  filter(between(year(londun_hefst), 2016, 2019)) |>
  collect(n = Inf) |>
  mutate(on.board = ifelse(londun_id %in% obs1$.id, TRUE, FALSE)) 
ln |>
  filter(veidarfaeri %in% c(6, 12, 3, 2, 13, 11, 9, 10, 7,
                            21, 9, 22)) |>
  group_by(londun_id, veidarfaeri, londun_hefst, on.board) |>
  mutate(on.board = ifelse(on.board, "Já", "Nei")) |>
  summarise(n.sid = n_distinct(fisktegund),
            .groups = "drop") |>
  mutate(dm = floor_date(londun_hefst, "month")) |>
  left_join(gear_code |> 
              rename(veidarfaeri = gid)) |> 
  mutate(veidarfaeri = paste(str_pad(veidarfaeri, 2, pad = "0"), gear)) |> 
  ggplot(aes(dm, n.sid, colour = on.board)) +
  stat_summary(fun.data = "mean_cl_boot",
               size = 0.25) +
  scale_colour_brewer(palette = "Set1") +
  facet_wrap(~ veidarfaeri, scales = "free_y") +
  labs(x = "Mánuður", y = "Fjöldi tegunda",
       colour = "Eftirlitsmaður",
       caption = "Hver rammi: Veiðarfæri, Miðgildi og 95% öryggismörk") +
  theme(legend.position = c(0.9, 0.15)) +
  expand_limits(y = 0)
```

