---
title: "Linking logbook, landings and ais"
warning: false
message: false
---

code on [github](https://github.com/einarhjorleifsson/fishydata)

like to comment: don't send e-mail, file an [issue](https://github.com/einarhjorleifsson/fishydata/issues)

* TODO:
* Add landings data link
* **Sanity checks**
* ...

## The data 

...

## Data compilation sequence

* [scripts/00_auxillary-data.R](https://github.com/einarhjorleifsson/fishydata/blob/master/scripts/00_auxillary-data.R)
* [scripts/01_vid-mid.R](https://github.com/einarhjorleifsson/fishydata/blob/master/scripts/scripts/01_vid-mid.R)
* [scripts/02-1_logbooks-merge.R](https://github.com/einarhjorleifsson/fishydata/blob/master/scripts/02-1_logbooks-merge.R)
* [scripts/02-2-logbooks-processing.R](https://github.com/einarhjorleifsson/fishydata/blob/master/scripts/02-2-logbooks-processing.R)
* [scripts/02-3_logbooks-for-ais.R](https://github.com/einarhjorleifsson/fishydata/blob/master/scripts/02-3_logbooks-for-ais.R)
* [scripts/04-stk-tails_parquet.R](https://github.com/einarhjorleifsson/fishydata/blob/master/scripts/04-stk-tails_parquet.R)

Should really set this up as a bash-script

Note: Log of the running kept in directory lgs

## Examples

```{r}
library(raster)
library(viridis)
library(leaflet)
library(sf)
library(tidyterra)
library(tidyverse)
library(arrow)
library(omar)
con <- connect_mar()
# data path:
pth <- "/net/hafkaldi.hafro.is/export/home/haf/einarhj/stasi/fishydata/data/"
## colour palettes -------------------------------------------------------------
hyp <- 
  tidyterra:::extract_pal(tidyterra::hypsometric_tints_db, 
                          palette = "etopo1_hypso") |> 
  pull(hex)
inf <- inferno(12, alpha = 1, begin = 0, end = 1, direction = -1)

# data -------------------------------------------------------------------------
island <- 
  geo::island |> 
  st_as_sf(coords = c("lon", "lat"),
           crs = 4326) |> 
  st_transform(crs = 3857) |> 
  st_coordinates() |> 
  as_tibble() |> 
  rename(x = X, y = Y)
sc_arrow <-
  tribble(~gid,  ~s1,  ~s2,
          -199, 1.000, 3.000,
          1,    0.375, 2.750,
          2,    0.125, 2.500,
          3,    0.025, 2.000,
          5,    0.250, 3.250,
          6,    2.625, 5.500,
          7,    2.625, 6.000,
          9,    2.375, 4.000,
          12,    0.050, 2.250,
          14,    1.750, 3.250,
          15,    0.500, 5.500,
          38,    0.250, 4.750,
          40,    0.250, 6.000) |> 
  mutate(gid = as.integer(gid)) |> 
  arrow::as_arrow_table()
# parquet files:
LB <- open_dataset(paste0(pth, "logbooks/station-for-ais.parquet"))
CT <- open_dataset(paste0(pth, "logbooks/catch-for-ais.parquet"))
con_ais <- 
  open_dataset(paste0(pth, "ais")) |> 
  left_join(sc_arrow)
# How many records:
con_ais |> count() |> pull()
```

### Relative effort

Here a simple relative effort of longline and bottom fishtrawl for all the years (2009-2022)

```{r}
#| fig.width: 9
#| fig.height: 9
#| fig.cap: "Relative efffort by gear based on 2009 to 2022 data."
dx <- 300
ais <-
  con_ais |> 
  mutate(year = year(time)) |> 
  filter(between(lon, -30, -20),
         between(lat, 65, 67.5),
         speed >= s1, speed <= s2,
         gid %in% c(1, 6)) |> 
  mutate(x = x %/% dx * dx + dx/2,
         y = y %/% dx * dx + dx/2,
         one = 1L) |> 
  group_by(gid, x, y) |> 
  summarise(z = sum(one),
            .groups = "drop") |> 
  collect()
ais |> 
  mutate(z = ifelse(z > quantile(z, 0.99),
                    quantile(z, 0.99),
                    z),
         gid = if_else(gid == 6, "Bottom trawl", "Longline", NA_character_)) |> 
  ggplot() +
  theme_void() +
  geom_tile(aes(x, y, fill = z)) +
  geom_polygon(data = island, aes(x, y),
               fill = "grey80") +
  tidyterra::scale_fill_hypso_c() +
  coord_equal(xlim = range(ais$x), ylim = range(ais$y)) +
  labs(fill = "Effort") +
  theme(plot.background = element_rect(fill = "grey", colour = "grey"),
        panel.background = element_rect(fill = "grey", colour = "grey"),
        plot.margin = unit(c(0,0,0,0), "cm")) +
  facet_wrap(~ gid, ncol = 1)
```

### Species catch

Catch distribution of Greenland halibut and Slope redfish for all years 2009-2022.

Done because of contemporaneous interest in bottom trawl closure off west Iceland.

```{r}
dx <- 300 # grid width
SID <- 61
# need to compute the pings per station id
con_ais_sid_count <- 
  con_ais |> 
  filter(speed >= s1, speed <= s2,
         gid == 6,
         lon >= -33, lon <= -10,
         lat >=  62, lat <= 68.5) |> 
  count(.sid) |> 
  compute()

catch61 <- 
  con_ais |> 
  filter(speed >= s1, speed <= s2,
         gid == 6,
         lon >= -33, lon <= -10,
         lat >=  62, lat <= 68.5) |> 
  left_join(con_ais_sid_count) |> 
  inner_join(CT |> 
               filter(sid == SID)) |> 
  mutate(catch = catch / n) |> 
  mutate(x = x %/% dx * dx + dx/2,
         y = y %/% dx * dx + dx/2) |> 
  group_by(x, y) |> 
  summarise(z = sum(catch) / 1e3,   # tonnes
            .groups = "drop") |> 
  collect()
SID <- 22
catch22 <- 
  con_ais |> 
  filter(speed >= s1, speed <= s2,
         gid == 6,
         lon >= -33, lon <= -10,
         lat >=  62, lat <= 68.5) |> 
  left_join(con_ais_sid_count) |> 
  inner_join(CT |> 
               filter(sid == SID)) |> 
  mutate(catch = catch / n) |> 
  mutate(x = x %/% dx * dx + dx/2,
         y = y %/% dx * dx + dx/2) |> 
  group_by(x, y) |> 
  summarise(z = sum(catch) / 1e3,   # tonnes
            .groups = "drop") |> 
  collect()
catch <-
  con_ais |> 
  filter(speed >= s1, speed <= s2,
         gid == 6,
         lon >= -33, lon <= -10,
         lat >=  62, lat <= 68.5) |> 
  left_join(con_ais_sid_count) |> 
  inner_join(CT |> 
               group_by(.sid) |> 
               summarise(catch = sum(catch, na.rm = TRUE),
                         .groups = "drop")) |> 
  mutate(catch = catch / n) |> 
  mutate(x = x %/% dx * dx + dx/2,
         y = y %/% dx * dx + dx/2) |> 
  group_by(x, y) |> 
  summarise(z = sum(catch) / 1e3,   # tonnes
            .groups = "drop") |> 
  collect()
p <- 
  bind_rows(catch22 |> mutate(sid = "G. halbut"),
            catch61 |> mutate(sid = "D. redfish")) |> 
  group_by(sid) |> 
  mutate(z = ifelse(z > quantile(z, 0.99),
                    quantile(z, 0.99),
                    z)) |> 
  ggplot() +
  theme_void() +
  geom_tile(aes(x, y, fill = z)) +
  geom_polygon(data = island, aes(x, y),
               fill = "grey80") +
  tidyterra::scale_fill_hypso_c() +
  coord_equal(xlim = range(ais$x), ylim = range(ais$y)) +
  labs(fill = "Catch") +
  theme(plot.background = element_rect(fill = "grey", colour = "grey"),
        panel.background = element_rect(fill = "grey", colour = "grey"),
        plot.margin = unit(c(0,0,0,0), "cm")) +
  facet_wrap(~ sid, ncol = 1)
p
```

```{r}
## area closures ---------------------------------------------------------------
closure <- 
  tribble(~lat, ~lon,
          655400, 274200,
          660600, 270000,
          654800, 270000,
          653000, 274200,
          655400, 274200) |> 
  mutate(lon = -geo::geoconvert.1(lon),
         lat =  geo::geoconvert.1(lat)) |> 
  st_as_sf(coords = c("lon", "lat"),
           crs = 4326) |> 
  st_transform(crs = 3857) |> 
  st_coordinates() |> 
  as_tibble() |> 
  rename(x = X, y = Y)
closure2 <-
  tribble(~lat, ~lon,
          660000, 270000,
          654300, 274200,
          653500, 274200,
          654300, 270000,
          660000, 270000) |> 
  mutate(lon = -geo::geoconvert.1(lon),
         lat =  geo::geoconvert.1(lat)) |> 
  st_as_sf(coords = c("lon", "lat"),
           crs = 4326) |> 
  st_transform(crs = 3857) |> 
  st_coordinates() |> 
  as_tibble() |> 
  rename(x = X, y = Y)
p +
  geom_path(data = closure,
            aes(x, y),
            colour = "red") +
  geom_path(data = closure2,
            aes(x, y),
            colour = "cyan") +
  coord_equal(xlim = c(-3.2e6, -2.8e6), ylim = c(9.5e6, 10e6))
```

```{r}
catch |> 
  filter(z > 0) |> 
  full_join(catch61 |> rename(z61 = z)) |> 
  mutate(z = replace_na(z, 0), # should not have this
         z61 = replace_na(z61, 0),
         r = z61 / z,
         r = ifelse(r > 1, 1, r),
         r = santoku::chop(r, breaks = c(0, 5, 10, 15, 25, 50, 75, 90, 100) / 100)) |> 
  ggplot() +
  theme_void() +
  geom_tile(aes(x, y, fill = r)) +
  geom_polygon(data = island, aes(x, y),
               fill = "grey80") +
  tidyterra::scale_fill_hypso_d() +
  coord_equal(xlim = c(-3.2e6, -2.8e6), ylim = c(9.5e6, 10e6)) +
  labs(fill = "p(Redfish") +
  theme(plot.background = element_rect(fill = "grey", colour = "grey"),
        panel.background = element_rect(fill = "grey", colour = "grey"),
        plot.margin = unit(c(0,0,0,0), "cm")) +
  geom_path(data = closure,
            aes(x, y),
            colour = "red") +
  geom_path(data = closure2,
            aes(x, y),
            colour = "cyan")
```

