---
title: "Trips explorations"
---

```{r base}
library(sf)
library(mapview)
library(tidyverse)
library(arrow)
source("R/read.R")
# parquet files:
pth = "/net/hafkaldi.hafro.is/export/home/haf/einarhj/stasi/fishydata/data/"
LB <- arrow::open_dataset(paste0(pth, "logbooks/station-for-ais.parquet"))
CT <- arrow::open_dataset(paste0(pth, "logbooks/catch-for-ais.parquet"))
AIS <- fd_stk_arrow() |> filter(vms == "yes")
AIS |> 
  filter(vms == "yes") |> 
  count(vid, .cid, .sid) |> 
  collect() |> 
  arrow::write_dataset("trip_visir.parquet")
TRIP_VISIR <- 
  arrow::open_dataset("trip_visir.parquet") |> 
  filter(!is.na(.sid))
LB <- 
  LB |> 
  left_join(TRIP_VISIR |> 
              select(vid, .sid, .cid)) |> 
  compute()
LB |> glimpse()

# create trips
AIS |> 
  filter(.cid > 0) |> 
  group_by(vid, .cid) |> 
  summarise(t1 = min(time),
            t2 = max(time),
            gid = min(gid, na.rm = TRUE),
            hid_dep = min(hid_dep),
            hid_arr = min(hid_arr),
            n_pings = n(),
            # what if we have NA's
            n_stations = n_distinct(.sid, na.rm = TRUE),
            .groups = "drop") |> 
  collect() |> 
  arrange(vid, t1) ->
  trips
# create stations
AIS |> 
  filter(.cid > 0) |> 
  group_by(vid, .cid, .sid) |> 
  summarise(t1 = min(time),
            t2 = max(time),
            gid = min(gid, na.rm = TRUE),
            n_pings = n(),
            .groups = "drop") |> 
  # get number of reported species
  inner_join(CT |> 
              group_by(.sid) |> 
              summarise(n_species = n_distinct(sid, na.rm = TRUE))) |> 
  collect() |> 
  arrange(vid, t1) ->
  stations
```

Should not have any trips with zero stations but gear assigned:
```{r gear-trips1}
trips |> 
  filter(n_stations == 0) |> 
  count(gid)
```

Should not have any trips with gear assigned but no stations:
```{r gear-trips2}
trips |> 
  filter(is.na(gid)) |> 
  count(n_stations)
```

### Distribution of stations per trip by gear

```{r stations-per-trip-by-year}
trips |> 
  filter(!is.na(gid)) |> 
  filter(gid %in% c(1:9, 14, 15, 18)) |> 
  mutate(n_stations = ifelse(n_stations > 40, 40, n_stations)) |> 
  count(gid, n_stations) |> 
  ggplot(aes(n_stations, n)) +
  geom_point() +
  facet_wrap(~ gid, scales = "free_y") +
  expand_limits(y = 0)
```

Unrealistic fishtrawl distribution:

```{r trawl-distributions}
trips |> 
  filter(!is.na(gid)) |> 
  filter(gid %in% c(6)) |> 
  #mutate(n_stations = ifelse(n_stations > 40, 40, n_stations)) |> 
  count(gid, n_stations) |> 
  ggplot(aes(n_stations, n)) +
  geom_point() +
  facet_wrap(~ gid, scales = "free_y") +
  expand_limits(y = 0)
trips |> 
  filter(gid == 6,
         n_stations == max(n_stations))
tmp <- 
  AIS |> 
  filter(vid == 2549,
         .cid == 41,
         vms == "yes") |> 
  collect() |> 
  st_as_sf(coords = c("lon", "lat"),
           crs = 4326)
mapview(tmp)
```

At least the max station number is assocated with a barent sea trip, but still not a completed trip.

### Number of species in a haul

```{r species-per-haul}
# let's select fishtrawl, 13 tows per trip
tmp <- 
  trips |> 
  filter(gid == 6,
         n_stations %in% 2:30) |> 
  select(vid, .cid, n_stations) |> 
  left_join(stations) |> 
  arrange(vid, t1) |> 
  group_by(vid, .cid, n_stations) |> 
  mutate(n_in_trip = 1:n()) |> 
  ungroup()
tmp.sum <-
  tmp |> 
  group_by(n_in_trip, n_stations) |> 
  summarise(m = mean(n_species, na.rm = TRUE),
            .groups = "drop")
tmp |> 
  ggplot() +
  theme_bw(base_size = 16) +
  geom_line(data = tmp.sum,
            aes(n_in_trip, m, group = n_stations,
                colour = factor(n_stations)),
            linewidth = 1) +
  #stat_summary(aes(n_in_trip, n_species, colour = factor(n_stations)),
  #             fun.data = "mean_cl_boot") +
  expand_limits(y = 0) +
  theme(legend.position = "none") +
  scale_colour_viridis_d() +
  labs(x = "Númer togs í túr",
       y = "Meðalfjöldi tegunda í togi",
       caption = "Fiskibotnvarpa - litir: hámarksfjöldi toga í túr") +
  scale_y_continuous(breaks = 1:10)
```

### Cumulative species numbers per trip

```{r cum-species-per-haul}
catch <- CT |> collect()

# for range of tows in trip
tmp <- 
  trips |> 
  filter(gid == 6,
         n_stations %in% 1:30) |> 
  select(vid, .cid, n_stations) |> 
  left_join(stations) |> 
  arrange(vid, t1) |> 
  group_by(vid, .cid, n_stations) |> 
  mutate(n_in_trip = 1:n()) |> 
  ungroup() |> 
  left_join(catch) |> 
  arrange(vid, t1, sid) |> 
  group_by(n_stations, vid, .cid, sid) |> 
  summarise(n_first = min(n_in_trip),
            .groups = "drop") |> 
  arrange(vid, .cid, n_first) |> 
  group_by(n_stations, vid, .cid, n_first) |> 
  summarise(csid = n()) |> 
  group_by(n_stations, vid, .cid) |> 
  mutate(csid = cumsum(csid)) |> 
  ungroup() 
tmp |> 
  group_by(n_stations, n_first) |> 
  summarise(csid = mean(csid)) |> 
  ggplot(aes(n_first, csid, colour = factor(n_stations))) +
  theme_bw(base_size = 16) +
  geom_point() +
  geom_line(linewidth = 1) +
  scale_colour_viridis_d() +
  theme(legend.position = "none") +
  expand_limits(y = 0) +
  scale_y_continuous(breaks = 0:14) +
  labs(x = "Númer togs í túr", y = "Uppsafnaður fjöldi tegunda",
       caption = "Fiskibotnvarpa - litir: hámarksfjöldi toga í túr")
```

What are the species added in the last tows per trip - by frequency:

```{r species-in-last-tow, fig.height = 9}
trips |> 
  filter(gid == 6,
         n_stations %in% 2:20) |> 
  select(vid, .cid, n_stations) |> 
  left_join(stations) |> 
  arrange(vid, t1) |> 
  group_by(vid, .cid, n_stations) |> 
  mutate(n_in_trip = 1:n()) |> 
  ungroup() |> 
  left_join(catch) |> 
  arrange(vid, t1, sid) |> 
  group_by(n_stations, vid, .cid, sid) |> 
  summarise(n_first = min(n_in_trip),
            .groups = "drop") |> 
  arrange(vid, .cid) |> 
  #filter(n_stations == 13) |> 
  filter(n_stations == n_first) |> 
  count(sid) |> 
  filter(sid != 184) |> 
  left_join(mar::les_tegund(mar::connect_mar()) |> 
              select(sid = tegund_nr, tegund = heiti) |> 
              collect(n = Inf)) |> 
  filter(!is.na(tegund)) |> 
  ungroup() |> 
  mutate(p = n / sum(n)) |> 
  ggplot(aes(p, reorder(tegund, p))) +
  geom_point() +
  labs(x = "Hlutfall", y = "Tegund") +
  scale_x_continuous(breaks = seq(0, 0.20, by = 0.02))
```


