---
title: "Fishtrawl footprint"
author: "Einar Hjörleifsson"
format: revealjs
echo: false
warning: false
message: false
---


```{r data}
library(sf)
library(mapdeck)
library(patchwork)
library(duckdbfs)
library(tidyverse)
ais <- open_dataset("/u3/geo/fishydata/data/ais/trail")
catch <- open_dataset("/u3/geo/fishydata/data/logbooks/catch-for-ais.parquet")
species <- 
  open_dataset("/u3/geo/fishydata/data/auxillary/agf_species.parquet")
dx <- 0.01
dy <- dx / 2



# 4 discrete colours from the turbo palette
cols <- viridisLite::turbo(4) |> rev()
cols2 <- c(RColorBrewer::brewer.pal(n = 12, "Paired"), "#63E6FF", "#0077B6", "#FF6FD8", "#A1006B")
cols2 <- c(RColorBrewer::brewer.pal(n = 12, "Paired"), "#4DDCCB", "#00796B", "#FF5BC8", "#8E005F")
# n <- 16
# image(1:n, 1, as.matrix(1:n), col = cols2, 
#          xlab = paste(cols2), ylab = "", xaxt = "n", yaxt = "n", 
#          bty = "n")

# Calculate the cumulatives - total catch --------------------------------------
aisg <- 
  ais |>
  filter(.cid > 0,
         gid_trip == 6,
         between(speed, s1, s2),
         between(time, t1, t2),
         between(lon, -30, -10),
         between(lat, 62.5, 69),
         between(year, 2009, 2024)) |>
  group_by(.sid, lb_base) |>
  mutate(catch = catch_total / n()) |>
  ungroup() |>
  mutate(x = lon%/%dx * dx + dx/2,
         y = lat%/%dy * dy + dy/2) |>
  group_by(x, y) |>
  summarise(catch = sum(catch, na.rm = TRUE),
            .groups = "drop") |> 
  filter(catch > 0) |> 
  collect() |> 
  arrange(catch) |> 
  mutate(n = 1:n(),
         pn = n / max(n, na.rm = TRUE),
         pc = catch / sum(catch, na.rm = TRUE),
         pc = cumsum(pc))
# Calculate the cumulatives - catch by species ---------------------------------
# Get major species
SIDS <- c(1:3, 5:9, 13, 14, 19, 22, 23, 24, 25, 61)
species <- c("Cod", "Haddock", "Saithe", "G. redfish", "Ling", "Blue ling", "Tusk",
             "Wolffish", "S. wolffish", "Monkfish", "Argentine", "G. halibut", "Plaice",
             "L. sole", "Witch", "D. redfish")
species <- 
  tibble(sid = SIDS) |> 
  mutate(species = species,
         hex = cols2)
res <- list()
for(i in 1:length(SIDS)) {
  sid_now <- SIDS[i]
  res[[i]] <-
    ais |>
    filter(.cid > 0,
         gid_trip == 6,
         between(speed, s1, s2),
         between(time, t1, t2),
         between(lon, -30, -10),
         between(lat, 62.5, 69),
         between(year, 2009, 2024)) |>
    inner_join(catch |>
                 filter(sid == sid_now),
               by = join_by(.sid, lb_base)) |>
    group_by(.sid, lb_base) |>
    mutate(catch = catch / n()) |>
    ungroup() |>
    mutate(x = lon%/%dx * dx + dx/2,
           y = lat%/%dy * dy + dy/2) |>
    group_by(x, y) |>
    summarise(catch = sum(catch, na.rm = TRUE),
              .groups = "drop") |>
    collect() |> 
    filter(catch > 0)
}
names(res) <- SIDS
aisg_species <- 
  bind_rows(res, .id = "sid") |> 
  mutate(sid = as.integer(sid)) |> 
  filter(catch > 0) |> 
  arrange(sid, catch) |>
  group_by(sid) |> 
  mutate(n = 1:n(),
         pn = n / max(n, na.rm = TRUE),
         pc = catch / sum(catch, na.rm = TRUE),
         pc = cumsum(pc)) |> 
  ungroup()
```

## Relative catch distribution

```{r relative_catch}

p_total_catch <- 
  aisg |> 
  mutate(p = case_when(between(pc, 0, 0.09999) ~ "00-10 %",
                       between(pc, 0.01, 0.25) ~ "10-25 %",
                       between(pc, 0.25, 0.50) ~ "25-50 %",
                       between(pc, 0.50, 1.00) ~ "50+ %",
                       .default = "something else")) |> 
  ggplot() +
  theme_void() +
  geom_polygon(data = geo::island, aes(lon, lat), fill = "grey") +
  geom_tile(aes(x, y, fill = p)) +
  coord_quickmap() +
  scale_fill_manual(values = cols) +
  theme(legend.position = c(0.6, 0.45)) +
  labs(fill = "Percentage of catch")
p_total_catch_90 <- 
  aisg |> 
  mutate(p = case_when(between(pc, 0, 0.09999) ~ "00-10 %",
                       between(pc, 0.01, 0.25) ~ "10-25 %",
                       between(pc, 0.25, 0.50) ~ "25-50 %",
                       between(pc, 0.50, 1.00) ~ "50+ %",
                       .default = "something else")) |> 
  ggplot() +
  theme_void() +
  geom_polygon(data = geo::island, aes(lon, lat), fill = "grey") +
  geom_tile(aes(x, y, fill = p)) +
  coord_quickmap() +
  scale_fill_manual(values = c("white", cols[2:4])) +
  labs(fill = "Percentage of catch") +
  theme(legend.position = "none")
p_total_catch
```

## Cumulative catch and footprint

```{r cumulative_catch}
p_cum <- 
  aisg |> 
  ggplot() +
  theme_bw(base_size = 15) +
  geom_line(aes(pc * 100, pn * 100),
            linewidth = 2) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10)) +
  labs(x = "Cumulative catch [%]", y = "Cumulative footprint [%]") +
  annotate("text", x = 2.5, y = 10, label = "Low catch\nin square", hjust = 0, size = 4) +
  annotate("text", x = 99, y = 10, label = "High catch\nin square", hjust = 1, size = 4)
p_cum
```

## Cumulative catch and footprint

```{r cumulative_catch_and_footprint}
r <- 
  tribble(~xmin, ~xmax, ~ymin, ~ymax, ~pc,
          0, 100, 0, 100,  "50+ %",
          0,  50, 0, 95, "25-50 %",
          0, 25,  0, 83, "10-25 %",
          0, 10, 0, 65, "00-10 %")
p_cum2 <- 
  aisg |> 
  ggplot() +
  theme_bw(base_size = 15) +
  geom_rect(data = r, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = pc),
            alpha = 0.75) +
  geom_line(aes(pc * 100, pn * 100),
            linewidth = 2) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10)) +
  labs(x = "Cumulative catch [%]", y = "Cumulative footprint [%]") +
  scale_fill_manual(values = cols) +
  theme(legend.position = "none")
p_cum2 + p_total_catch
```

## 90% of total catch 

```{r 90_percent_of_catch}
p_total_catch + 
  labs(subtitle = "All catch") +
  p_total_catch_90 +
  labs(subtitle = "90% of catch")
```

## What about CPUE?

```{r cpue}
cpue <- 
  ais |>
  filter(.cid > 0,
         whack != TRUE,
         gid_trip == 6,
         between(speed, s1, s2),
         between(time, t1, t2),
         between(lon, -30, -10),
         between(lat, 62.5, 69),
         between(year, 2009, 2024)) |>
  group_by(.sid, lb_base) |>
  mutate(catch = catch_total / n()) |>
  ungroup() |>
  mutate(x = lon%/%dx * dx + dx/2,
         y = lat%/%dy * dy + dy/2) |>
  group_by(x, y) |>
  summarise(catch = sum(catch, na.rm = TRUE) / 1000,
            effort = sum(dt, na.rm = TRUE) / (60 * 60),
            .groups = "drop") |> 
  filter(catch > 0, 
         effort > 0) |> 
  collect()
cpue |>
  mutate(cpue = catch / effort) |> 
  #mutate(cpue = ifelse(cpue > 3, 3, cpue)) |> 
  mutate(cpue = case_when(between(cpue, 0, 0.9) ~ "0000-0900",
                          #between(cpue, 0.1, 0.5) ~ "0100-0500",
                          between(cpue, 0.9, 1.5) ~ "0900-1500",
                          between(cpue, 1.5, 2.5) ~ "1500-2500",
                          between(cpue, 2.5, Inf) ~ "2500+",
                          .default = "Something else")) |> 
  #ggplot() + geom_histogram(aes(effort))
  ggplot() +
  theme_void() +
  geom_polygon(data = geo::island,
               aes(lon, lat),
               fill = "grey") +
  geom_tile(aes(x, y, fill = cpue)) +
  scale_fill_viridis_d(option = "turbo", direction = -1) +
  coord_quickmap() +
  labs(fill = "CPUE [kg/hour]")
```


## 8 species

```{r 8_species}
p <-
  aisg_species |>
  filter(sid %in% c(1, 2, 3, 5, 6, 7, 8, 9)) |> 
  left_join(species) |> 
  mutate(sid = paste(str_pad(sid, width = 2, pad = "0"), species)) |> 
  mutate(p = case_when(between(pc, 0, 0.09999) ~ "00-10 %",
                       between(pc, 0.01, 0.25) ~ "10-25 %",
                       between(pc, 0.25, 0.50) ~ "25-50 %",
                       between(pc, 0.50, 1.00) ~ "50+ %",
                       .default = "something else")) |> 
  ggplot() +
  theme_void() +
  geom_polygon(data = geo::island, aes(lon, lat), fill = "grey") +
  geom_tile(aes(x, y, fill = p)) +
  coord_quickmap() +
  theme(legend.position = c(0.5, 0.5)) +
  labs(fill = "Percentage of catch") +
  facet_wrap(~ sid, ncol = 4)
p + scale_fill_manual(values = cols)
```

## 8 species - 90% of catch

```{r}
p + scale_fill_manual(values = c("white", cols[2:4]))
```

## 8 species - 90% of catch

```{r}
aisg_species |>
  filter(pc >= 0.1) |> 
  filter(sid %in% c(1, 2, 3, 5, 6, 7, 8, 9)) |> 
  left_join(species) |> 
  mutate(sid = paste(str_pad(sid, width = 2, pad = "0"), species)) |> 
  mutate(p = case_when(between(pc, 0, 0.09999) ~ "00-10 %",
                       between(pc, 0.01, 0.25) ~ "10-25 %",
                       between(pc, 0.25, 0.50) ~ "25-50 %",
                       between(pc, 0.50, 1.00) ~ "50+ %",
                       .default = "something else")) |> 
  arrange(desc(pc), sid) |> 
  group_by(x, y) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot() +
  theme_void() +
  geom_polygon(data = geo::island, aes(lon, lat), fill = "grey") +
  geom_tile(aes(x, y, fill = sid)) +
  coord_quickmap() +
  ggmisc::scale_fill_crayola(guide = "legend") +
  scale_fill_manual(values = cols2) +
  labs(fill = "Species") +
  theme(legend.position = c(0.65, 0.45))
```

## 8 more species

```{r 8_species2}
p <-
  aisg_species |>
  filter(!sid %in% c(1, 2, 3, 5, 6, 7, 8, 9)) |> 
  left_join(species) |> 
  mutate(sid = paste(str_pad(sid, width = 2, pad = "0"), species)) |> 
  mutate(p = case_when(between(pc, 0, 0.09999) ~ "00-10 %",
                       between(pc, 0.01, 0.25) ~ "10-25 %",
                       between(pc, 0.25, 0.50) ~ "25-50 %",
                       between(pc, 0.50, 1.00) ~ "50+ %",
                       .default = "something else")) |> 
  ggplot() +
  theme_void() +
  geom_polygon(data = geo::island, aes(lon, lat), fill = "grey") +
  geom_tile(aes(x, y, fill = p)) +
  coord_quickmap() +
  theme(legend.position = c(0.5, 0.5)) +
  labs(fill = "Percentage of catch") +
  facet_wrap(~ sid, ncol = 4)
p + scale_fill_manual(values = cols)
```

## 8 more species - 90% of catch

```{r}
p + scale_fill_manual(values = c("white", cols[2:4]))
```

## 8 more species - 90% of catch

```{r}
aisg_species |>
  filter(pc >= 0.1) |> 
  filter(!sid %in% c(1, 2, 3, 5, 6, 7, 8, 9)) |> 
  left_join(species) |> 
  mutate(sid = paste(str_pad(sid, width = 2, pad = "0"), species)) |> 
  mutate(p = case_when(between(pc, 0, 0.09999) ~ "00-10 %",
                       between(pc, 0.01, 0.25) ~ "10-25 %",
                       between(pc, 0.25, 0.50) ~ "25-50 %",
                       between(pc, 0.50, 1.00) ~ "50+ %",
                       .default = "something else")) |> 
  arrange(desc(pc), sid) |> 
  group_by(x, y) |> 
  slice(1) |> 
  ungroup() |> 
  ggplot() +
  theme_void() +
  geom_polygon(data = geo::island, aes(lon, lat), fill = "grey") +
  geom_tile(aes(x, y, fill = sid)) +
  coord_quickmap() +
  #ggmisc::scale_fill_crayola(guide = "legend") +
  scale_fill_manual(values = cols2[9:16]) +
  labs(fill = "Species") +
  theme(legend.position = c(0.65, 0.45))
```


## 16 species - 90% of catch

```{r}
aisg_species |>
  filter(pc >= 0.1) |> 
  arrange(desc(pc), sid) |> 
  group_by(x, y) |> 
  slice(1) |> 
  ungroup() |> 
  left_join(species) |>
  mutate(sid = paste(str_pad(sid, width = 2, pad = "0"), species)) |> 
  ggplot() +
  theme_void() +
  geom_polygon(data = geo::island, aes(lon, lat), fill = "grey") +
  geom_tile(data = aisg, aes(x, y), fill = "grey") +
  geom_tile(aes(x, y, fill = as.factor(sid))) +
  coord_quickmap() +
  scale_fill_manual(values = cols2) +
  labs(fill = "Species")
```

## 16 species - 90% of catch

```{r}
m <- 
  aisg_species |>
  filter(pc >= 0.1) |> 
  arrange(desc(pc), sid) |> 
  group_by(x, y) |> 
  slice(1) |> 
  ungroup() |> 
  left_join(species) |>
  mutate(sid = paste(str_pad(sid, width = 2, pad = "0"), species)) |> 
  select(lon = x, lat = y, speed = sid, hex) |> 
  sf::st_as_sf(coords = c("lon", "lat"), crs = 4326)
m2 <-
  m |> 
  mutate(hex = paste0(hex, "FF"))
mapdeck(m2, style = mapdeck_style("dark")) |> 
  mapdeck::add_scatterplot(tooltip = "speed", 
                           radius = 350,
                           colour = "hex",
                           fill_colour = "hex",
                           layer_id = "points")
```


## Catch 22


* Use 0.15 or 0.30 minute interpolation of pings
* Area of each square is a function of latitude - need to correct for that
* Check for outliers - cap large catches
* Think about less confusing statistics


Geheimnis:

```{r}
footprint <- aisg |> select(x, y) |> distinct()
footprint |>
  left_join(aisg_species |>
              ungroup() |> 
              filter(pc >= 0.1) |> 
              select(x, y) |> 
              distinct() |> 
              mutate(what = "open")) |>
  mutate(what = replace_na(what, "closed")) |>
  count(what) |>
  mutate(p = n / sum(n))
```
