---
title: "Working with AIS data using {duckdb} and {h3}"
author: Einar Hjörleifsson
description: "Must admit that I have not fully grasped the duckdb-concept but what I have explored is magical. The main R-package is {duckdb} but {duckdbfs} provides a nice wrapper for making connection to files easier (similar to the {arrow}) package. On top of it {duckdbfs} has some intriguing spatial function. Here a little exploration of the AIS data is made to put the feature to a test."
date: "2025-06-15"
categories: [code, rtip]
echo: true
---

## The steps

* Open a duckdb connection to a sweep of AIS parquet files.
  * Datafiles contain all Icelandic vessel pings from 2008-2024, total of ~475 million records!
* Filter the data, only pings of demersal fish bottom trawl, limit the extent to fishing in Icelandic waters.
* Convert coordinates to h3-index at resolution level 9. This means that each hexagonal grid is 0.105	$km^2$ (If this were a circle the diameter would be ~400 meters)
* Sum effort for each hex-grid for each year
* Above all done within duckdb, the data only imported into R in the last step

This results in ~7.4 million records consisting of three variables (year, hex-grid index and effort), the time the whole processing takes is less 10 seconds!

## The raw data structure

The actual dataset (475 million records) structure om my computer is something like this (each year of data is a separate hive-partioned parquet file):

```
rail
├── year=2007
│   └── part-0.parquet
├── year=2008
│   └── part-0.parquet
├── year=2009
│   └── part-0.parquet
├── year=2010
│   └── part-0.parquet
├── year=2011
│   └── part-0.parquet
├── year=2012
│   └── part-0.parquet
├── year=2013
│   └── part-0.parquet
├── year=2014
│   └── part-0.parquet
├── year=2015
│   └── part-0.parquet
├── year=2016
│   └── part-0.parquet
├── year=2017
│   └── part-0.parquet
├── year=2018
│   └── part-0.parquet
├── year=2019
│   └── part-0.parquet
├── year=2020
│   └── part-0.parquet
├── year=2021
│   └── part-0.parquet
├── year=2022
│   └── part-0.parquet
├── year=2023
│   └── part-0.parquet
├── year=2024
│   └── part-0.parquet
└── year=2025
    └── part-0.parquet
```


## The processing

```{r}
library(tidyverse)
library(duckdbfs)
library(mapdeck)
library(tictoc)
tic()
zoom <- 9L
tic()
q <-
  open_dataset(here::here("data/ais/trail"))
# peek-a-boo 
q |> glimpse()
d <- 
  q |> 
  filter(between(lon, -30, -10),        # ~ Icelandic area
         between(lat, -62.5, 68),
         between(speed, s1, s2),        # only towing speed
         between(time, t1, t2),         # start-end of tow
         .cid > 0,                      # not in harbour
         agf_gid == 6) |>               # bottom trawl
  mutate(h3id = h3_latlng_to_cell_string(lat, lon, zoom)) |> 
  group_by(year, h3id) |> 
  summarise(z = sum(dt, na.rm = TRUE),  # sum the step time
            .groups = "drop") |> 
  filter(!is.na(z)) |> 
  # only now do we import the data into R
  collect()         
toc()
d |> glimpse()
```

## A little map

We then use the 'add_h3' {mapdeck}-function to get a quick overview of the results, here only for year 2024.

```{r}
mapdeck(location = c(-22, 66),
        zoom = 6,
        libraries = "h3") |> 
  add_h3(data = 
           d |> 
           filter(year == 2024) |> 
           # trim the extremes
           mutate(z = ifelse(z > quantile(z, 0.99), quantile(z, 0.99), z)),
         hexagon = "h3id",
         fill_colour = "z",
         palette = "inferno")
```

