---
title: "GLM notes"
echo: FALSE
---

```{r}
library(arrow)
library(tidyverse)
library(ggfortify)
library(patchwork)
library(broom)
library(ciTools)
library(modelr)
# landings ---------------------------------------------------------------------
gear <-  
  read_rds(here::here("data/landings/agf_gear.rds")) |> 
  select(gid_ln = gid, heiti)
ob <- 
  read_rds(here::here("data/landings/agf_observers.rds")) |> 
  filter(use) |> 
  select(.lid) |> 
  mutate(observer = "yes")
CT <- read_rds(here::here("data/landings/agf_catch.rds"))
ca <- 
  CT |> 
  group_by(.lid) |> 
  summarise(n_sid = n_distinct(sid))
ln <- 
  read_rds(here::here("data/landings/agf_stations.rds")) |> 
  left_join(ca) |> 
  left_join(ob) |> 
  left_join(gear) |> 
  mutate(observer = replace_na(observer, "no"),
         year = year(datel),
         gear = paste0(str_pad(gid_ln, width = 2, pad = "0"),
                       " ",
                       heiti))
# # logbooks ---------------------------------------------------------------------
# pth = "/net/hafkaldi.hafro.is/export/home/haf/einarhj/stasi/fishydata/data/"
# lb_species <- 
#   arrow::open_dataset(paste0(pth, "logbooks/catch-for-ais.parquet")) |> 
#   group_by(.sid) |> 
#   summarise(n.species = n_distinct(sid)) |> 
#   collect()
# TRIP_VISIR <- 
#   arrow::open_dataset("trip_visir.parquet") |> 
#   filter(!is.na(.sid))
# lb <- 
#   arrow::open_dataset(paste0(pth, "logbooks/station-for-ais.parquet")) |> 
#   left_join(TRIP_VISIR |> 
#               select(vid, .sid, .cid)) |> 
#   collect() |> 
#   left_join(lb_species) |> 
#   arrange(vid, date, t1) |> 
#   group_by(vid, .cid) |> 
#   mutate(tow.n = 1:n(),
#          n.stations = max(tow.n)) |> 
#   ungroup()
species <- 
  omar::tbl_mar(omar::connect_mar(), "agf.aflagrunnur_v") |> 
  select(starts_with("fisktegund")) |> 
  distinct() |> 
  collect() |> 
  select(sid = fisktegund, tegund = fisktegund_heiti)

# # auxillary --------------------------------------------------------------------
# vids <- 
#   read_rds("data-aux/vessels.rds") |> 
#   select(vid, length)

# filters ----------------------------------------------------------------------
ln <- 
  ln |> 
  filter(year >= 2016,
         gid_ln %in% c(2, 3, 6, 7, 9, 11, 12, 13, 21)) |> 
  mutate(date = floor_date(datel, "month"),
         Eftirlitsmaður = ifelse(observer == "yes", "Já", "Nei"))


```


## Simple case

* Let's limit our analsys to  trawlers where observer have been on-board at least once
* Let's take a very simple case of only two trawlers, selected because they are:
  * At the extreme range of the number of species landed
  * Have some few trips where observer was onboard

```{r}
obs.vids <- 
  ln |> 
  filter(gid_ln == 6,
         Eftirlitsmaður == "Já") |> 
  pull(vid) |> 
  unique()
d <-
  ln |> 
  filter(gid_ln == 6,
         vid %in% obs.vids)

d2 <- 
  d |> 
  filter(vid %in% c(1937, 2170)) |> 
  select(vid, sid = n_sid, observer = Eftirlitsmaður) |> 
  mutate(vid = as.factor(vid),
         observer = as.factor(observer))
```

A simple histogram of the observations ...

```{r}
p <-
  d2 |> 
  count(vid, sid) |> 
  ggplot() +
  theme_bw() +
  geom_col(aes(sid, n),
           fill = "grey") +
  facet_grid(vid ~ ., scales = "free_y") +
  labs(x = "Number of species",
       y = "Number of trips")
p
```

#### Fit a GLM to the data

Here let's start by using gaussian-distribution (read: normal-distribution) just because it is simple to interpret the parameter estimates. A statistician may argue that we should use the poisson-distribution because we are dealing with counts (and we can not have negative count). Need to check that at some stage.

```{r}
m <- glm(formula = sid  ~ vid, data = d2, family = gaussian)
m |> 
  broom::glance() |> 
  knitr::kable(digits = 1)
m |> 
  broom::tidy(conf.int = TRUE) |> 
  knitr::kable(digits = 2)
autoplot(m)
```


* The estimated mean number of species for the first vessel (vid = 1937) is 8.55 species, with a 95% confidence interval between 8.31-8.80. This vessel servers as our ground-zero when interpreting the meaning of the parameter values of the second vessel.
* For vessel 2170 we have a value of 6.45 with a confidence interval of 5.89-701. The value of the estimate here is actually relative to the value of the first vessel (vid = 1937), ergo the mean number of species is 8.55 + 6.45 = 15.00

In order to get the actual mean estimate of the number of species for the second vessel we could do something like:


```{r}
pred <- 
  d2 |> 
  add_ci(m, alpha = 0.1) %>%
  select(vid, pred:UCB0.95) |> 
  distinct() |> 
  janitor::clean_names() 
pred |> 
  knitr::kable(digits = 2)
p +
  geom_rect(data = pred,
            aes(xmin =  lcb0_05,
                xmax = ucb0_95,
                ymin = -Inf,
                ymax =  Inf),
            fill = "red",
            alpha = 0.5) +
  geom_vline(data = pred, 
             aes(xintercept = pred),
             colour = "blue")
```

## Considering observers

Looking at the distribution adding the observer we have:

```{r}
p2 <- 
  d2 |>
  count(vid, observer, sid) |> 
  ggplot() +
  theme_bw() +
  geom_col(aes(sid, n, fill = observer)) +
  facet_grid(vid ~ ., scales = "free_y") +
  labs(x = "Number of species",
       y = "Number of trips") +
  scale_fill_brewer(palette = "Set1")
p2
```

We observe:

* The number of observer trips is a small proportion of the total trips
* When observer is on board the number of species landed tend to be in the uppper range of the distribution

Could get a quick summary view of the mean and the confidence interval via:

```{r}
d2 |> 
  ggplot(aes(vid, sid, colour = observer)) +
  stat_summary(fun.data = "mean_cl_boot") +
  scale_colour_brewer(palette = "Set1") +
  labs(x = "Vessel",
       y = "Number of species") +
  scale_y_continuous(breaks = 8:20)
```

So, a quick analysis indicates that we have significantly higher number of species landed when observer is on-board. Let's estimate the parameters:

```{r}
m <- glm(formula = sid  ~ vid + observer, data = d2, family = gaussian)
m |> 
  broom::glance() |> 
  knitr::kable(digits = 1)
m |> 
  broom::tidy(conf.int = TRUE) |> 
  knitr::kable(digits = 2)
autoplot(m)
```

So again the statistics are significant, both the vessel factor and the observer factor. Take note that the intercept is now different, 8.55 (species), is now  12.62. 

The meaning of 12.61 can be considered as the expected mean number of species landed by the first vessel when the observer is on-board, taken into account the estimated mean difference in the number of species (based on both vessels).

```{r}
pred <- 
  d2 |> 
  add_ci(m, alpha = 0.1) %>%
  select(vid, observer, pred:UCB0.95) |> 
  distinct() |> 
  janitor::clean_names() 
pred |> 
  knitr::kable(digits = 2)
```


#### Adding quarter

```{r}
d2 <- 
  d |> 
  mutate(quarter = quarter(date)) |> 
  filter(vid %in% c(1937, 2170)) |> 
  select(vid, sid = n_sid, q = quarter, observer = Eftirlitsmaður) |> 
  mutate(vid = as.factor(vid),
         q = as.factor(q),
         observer = as.factor(observer))
d2 |> count(vid, q, observer) |> spread(observer, n)
d2 |> 
  ggplot(aes(q, sid, colour = observer)) +
  stat_summary(fun.data = "mean_cl_boot") +
  facet_grid(vid ~ .) +
  scale_colour_brewer(palette = "Set1")
```

```{r}
m <- glm(formula = sid  ~ vid + observer + q, data = d2, family = gaussian)
m |> 
  broom::glance() |> 
  knitr::kable(digits = 1)
m |> 
  broom::tidy(conf.int = TRUE) |> 
  knitr::kable(digits = 2)
autoplot(m)
```

So adding quarter is not significant (so in a stepwise model this model should not be part of the final model).

#### Adding harbour

A bit meaningless

```{r}
d2 <- 
  d |> 
  filter(vid %in% c(1937, 2170)) |> 
  select(vid, sid = n_sid, hid = hid_ln, observer = Eftirlitsmaður) |> 
  mutate(vid = as.factor(vid),
         hid = as.factor(hid),
         observer = as.factor(observer))
d2 |> count(vid, hid, observer) |> spread(observer, n)
d2 |> 
  ggplot(aes(hid, sid, colour = observer)) +
  stat_summary(fun.data = "mean_cl_boot") +
  facet_grid(vid ~ .) +
  scale_colour_brewer(palette = "Set1")
```

```{r}
m <- glm(formula = sid  ~ vid + observer + hid, data = d2, family = gaussian)
m |> 
  broom::glance() |> 
  knitr::kable(digits = 1)
m |> 
  broom::tidy(conf.int = TRUE) |> 
  knitr::kable(digits = 2)
```


### The full dataset

```{r}
d <- 
  ln |> 
  mutate(year = year(datel),
         month = month(datel)) |> 
  filter(gid_ln == 6,
         vid %in% obs.vids) |> 
  select(sid = n_sid, vid, hid = hid_ln, month, observer = Eftirlitsmaður) |> 
  mutate(vid = factor(vid),
         hid = factor(hid),
         month = factor(month),
         observer = factor(observer))
m <- glm(formula = sid  ~ vid + observer + vid, data = d, family = gaussian)
m |> 
  broom::glance() |> 
  knitr::kable(digits = 1)
# m |> 
#   broom::tidy(conf.int = TRUE) |> 
#   knitr::kable(digits = 2)
autoplot(m)

pred <- 
  d |> 
  add_ci(m, alpha = 0.1) %>%
  select(vid, observer, pred:UCB0.95) |> 
  distinct() |> 
  janitor::clean_names() 
pred |> 
  ggplot() +
  geom_pointrange(aes(reorder(vid, pred),
                      y = pred,
                      ymin = lcb0_05, 
                      ymax = ucb0_95,
                      colour = observer)) +
  scale_colour_brewer(palette = "Set1") +
  scale_y_continuous(breaks = 0:25) +
  labs(x = "Vessel",
       y = "Number of species landed") +
  coord_flip()
```

Notes for the statistician:

* Error bar the same for no observer vs observer
* Constant difference in the number of species - need an interactive term?

### Add interactive term

```{r}
m <- glm(formula = sid  ~ vid + observer + vid * observer, data = d, family = gaussian)
m |> 
  broom::glance() |> 
  knitr::kable(digits = 1)
# m |> 
#   broom::tidy(conf.int = TRUE) |> 
#   knitr::kable(digits = 2)
autoplot(m)

pred <- 
  d |> 
  add_ci(m, alpha = 0.1) %>%
  select(vid, observer, pred:UCB0.95) |> 
  distinct() |> 
  janitor::clean_names() 
pred |> 
  ggplot() +
  geom_pointrange(aes(reorder(vid, pred),
                      y = pred,
                      ymin = lcb0_05, 
                      ymax = ucb0_95,
                      colour = observer)) +
  scale_colour_brewer(palette = "Set1") +
  scale_y_continuous(breaks = 0:25) +
  labs(x = "Vessel",
       y = "Number of species landed") +
  coord_flip()
```

### Try poisson (because these are counts)

```{r}
m <- glm(formula = sid  ~ vid + observer + vid * observer, data = d, family = poisson)
m |> 
  broom::glance() |> 
  knitr::kable(digits = 1)
# m |> 
#   broom::tidy(conf.int = TRUE) |> 
#   knitr::kable(digits = 2)
autoplot(m)

pred <- 
  d |> 
  add_ci(m, alpha = 0.1) %>%
  select(vid, observer, pred:UCB0.95) |> 
  distinct() |> 
  janitor::clean_names() 
pred |> 
  ggplot() +
  geom_pointrange(aes(reorder(vid, pred),
                      y = pred,
                      ymin = lcb0_05, 
                      ymax = ucb0_95,
                      colour = observer)) +
  scale_colour_brewer(palette = "Set1") +
  scale_y_continuous(breaks = 0:25) +
  labs(x = "Vessel",
       y = "Number of species landed") +
  coord_flip()
```

