---
title: "Landings"
---

```{r}
library(arrow)
library(tidyverse)
library(patchwork)
library(broom)
library(modelr)
# landings ---------------------------------------------------------------------
gear <-  
  read_rds(here::here("data/landings/agf_gear.rds")) |> 
  select(gid_ln = gid, heiti)
ob <- 
  read_rds(here::here("data/landings/agf_observers.rds")) |> 
  filter(use) |> 
  select(.lid) |> 
  mutate(observer = "yes")
CT <- read_rds(here::here("data/landings/agf_catch.rds"))
ca <- 
  CT |> 
  group_by(.lid) |> 
  summarise(n_sid = n_distinct(sid))
ln <- 
  read_rds(here::here("data/landings/agf_stations.rds")) |> 
  left_join(ca) |> 
  left_join(ob) |> 
  left_join(gear) |> 
  mutate(observer = replace_na(observer, "no"),
         year = year(datel),
         gear = paste0(str_pad(gid_ln, width = 2, pad = "0"),
                       " ",
                       heiti))
# # logbooks ---------------------------------------------------------------------
# pth = "/net/hafkaldi.hafro.is/export/home/haf/einarhj/stasi/fishydata/data/"
# lb_species <- 
#   arrow::open_dataset(paste0(pth, "logbooks/catch-for-ais.parquet")) |> 
#   group_by(.sid) |> 
#   summarise(n.species = n_distinct(sid)) |> 
#   collect()
# TRIP_VISIR <- 
#   arrow::open_dataset("trip_visir.parquet") |> 
#   filter(!is.na(.sid))
# lb <- 
#   arrow::open_dataset(paste0(pth, "logbooks/station-for-ais.parquet")) |> 
#   left_join(TRIP_VISIR |> 
#               select(vid, .sid, .cid)) |> 
#   collect() |> 
#   left_join(lb_species) |> 
#   arrange(vid, date, t1) |> 
#   group_by(vid, .cid) |> 
#   mutate(tow.n = 1:n(),
#          n.stations = max(tow.n)) |> 
#   ungroup()
species <- 
  omar::tbl_mar(omar::connect_mar(), "agf.aflagrunnur_v") |> 
  select(starts_with("fisktegund")) |> 
  distinct() |> 
  collect() |> 
  select(sid = fisktegund, tegund = fisktegund_heiti)

# # auxillary --------------------------------------------------------------------
# vids <- 
#   read_rds("data-aux/vessels.rds") |> 
#   select(vid, length)

# filters ----------------------------------------------------------------------
ln <- 
  ln |> 
  filter(year >= 2016,
         gid_ln %in% c(2, 3, 6, 7, 9, 11, 12, 13, 21)) |> 
  mutate(date = floor_date(datel, "month"),
         Eftirlitsmaður = ifelse(observer == "yes", "Já", "Nei"))

species_hz <- function(GID, ln,  CT, trim = TRUE) {
  trips <- 
    ln |> 
    filter(gid_ln == GID)
  GEAR <- trips |> pull(gear) |> unique()
  trips <-
    trips |> 
    count(Eftirlitsmaður)
  tmp <- 
    ln |> 
    filter(gid_ln == GID) |> 
    left_join(CT) |> 
    select(.lid, Eftirlitsmaður, sid) |> 
    distinct() |> 
    group_by(Eftirlitsmaður, sid) |> 
    summarise(n.species = n(),
              .groups = "drop") |> 
    left_join(trips) |> 
    left_join(species) |> 
    mutate(sid = paste0(str_pad(sid, width = 3, pad = "0"),
                        " ",
                        tegund)) |> 
    mutate(p.species = n.species / n)
  
  if(trim) {
    sids <- 
      tmp |> 
      select(Eftirlitsmaður, sid, p.species) |> 
      spread(Eftirlitsmaður, p.species, fill = 0) |> 
      arrange(Já) |> 
      filter(Já >= 0.001 | Nei >= 0.001) |> 
      pull(sid)
    tmp <- 
      tmp |> 
      filter(sid %in% sids)
  }
  
  p <- 
    tmp |> 
    ggplot() +
    geom_pointrange(aes(x = sid,
                        y = p.species,
                        ymin = 0,
                        ymax = p.species,
                        colour = Eftirlitsmaður)) +
    scale_colour_brewer(palette = "Set1") +
    scale_y_continuous(breaks = seq(0, 1, by = 0.1)) +
    labs(y = "Hlutfall landana",
         x = NULL,
         caption = GEAR) +
    theme(legend.position = c(0.8, 0.8)) +
    coord_cartesian(ylim = c(0, 1)) +
    coord_flip()
  return(p)
}
# Frequency of species ---------------------------------------------------------

if(FALSE) {
  species_hz(2, ln, CT)
  species_hz(3, ln, CT)
  species_hz(6, ln, CT)
  species_hz(7, ln, CT)
  species_hz(9, ln, CT)
  species_hz(11, ln, CT)
  species_hz(12, ln, CT)
  species_hz(13, ln, CT)
  species_hz(21, ln, CT)
}

trips <- 
  ln |> 
  count(Eftirlitsmaður, gear)

tmp <- 
  ln |> 
  left_join(CT) |> 
  select(.lid, Eftirlitsmaður, sid, gear) |> 
  distinct() |> 
  group_by(Eftirlitsmaður, sid, gear) |> 
  summarise(n.species = n(),
            .groups = "drop") |> 
  left_join(trips) |> 
  left_join(species) |> 
  mutate(sid = paste0(str_pad(sid, width = 3, pad = "0"),
                      " ",
                      tegund)) |> 
  mutate(p.species = n.species / n)

what <- 
  tmp |> 
  select(Eftirlitsmaður, sid, p.species, gear) |> 
  spread(Eftirlitsmaður, p.species, fill = 0) |> 
  arrange(Já) |> 
  filter(Já >= 0.001 | Nei >= 0.001) |> 
  select(sid, gear) |> 
  distinct()
tmp <- 
  tmp |> 
  inner_join(what)
```


```{r}
#| fig.height: 30
tmp |> 
  ggplot() + 
  geom_pointrange(aes(x = sid,
                      y = p.species,
                      ymin = 0,
                      ymax = p.species,
                      colour = Eftirlitsmaður)) +
  facet_grid(gear ~ ., scales = "free", space = "free") +
  theme(strip.text.y = element_text(angle = 0)) +
  labs(x = NULL, y = "Hlutfall landana") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1)) +
  coord_flip() +
  scale_colour_brewer(palette = "Set1")

```

## Viðauki

### Understanding general linear models 

* Let's limit our analsys to  trawlers where observer have been on-board
* Let's take a very simple case of only two trawlers, selected because they are:
  * At the extreme range of the number of species landed
  * Have some few trips where observer was onboard

```{r}
obs.vids <- 
  ln |> 
  filter(gid_ln == 6,
         Eftirlitsmaður == "Já") |> 
  pull(vid) |> 
  unique()
d <-
  ln |> 
  filter(gid_ln == 6,
         vid %in% obs.vids)

d2 <- 
  d |> 
  filter(vid %in% c(1937, 2170)) |> 
  select(vid, sid = n_sid, observer = Eftirlitsmaður) |> 
  mutate(vid = as.factor(vid),
         observer = as.factor(observer))
```

#### A simple histogram of the species per trip

```{r}
d2 |> 
  count(vid, sid) |> 
  ggplot() +
  geom_col(aes(sid, n)) +
  facet_grid(vid ~ ., scales = "free_y") +
  labs(x = "Number of species",
       y = "Number of trips")
```

#### Fit a GLM to the data

Here let's start by using gaussian-distribution (read: normal-distribution) just because it is simple to interpret the parameter estimates. A statistician may argue that we should use the poisson-distribution because we are dealing with counts (and we can not have negative count). Need to check that at some stage.

```{r}
m <- glm(formula = sid  ~ vid, data = d2, family = gaussian)
summary(m)
```

The statistics indicate that the model is highly significant. The meaning of the parameters is as follows:

```{r}
m |> 
  tidy(conf.int = TRUE) |> 
  knitr::kable(digits = 2)
```

* The estimated mean number of species for the first vessel (vid = 1937) is 8.55 species, with a 95% confidence interval between 8.43-8.68. This vessel servers as our ground-zero when interpreting the meaning of the parameter values of the second vessel.
* For vessel 2170 we have a value of 6.45 with a confidence interval of 6.16-6.73. This is to be interpreted as this vessel on average reportin 6.45 more species than the first vessel.

In order to get the actual mean estimate of the number of species for the second vessel we could do something like:

```{r}
m2179 <- 8.55 + 6.45   # the mean
```

```{r}
d2 |> 
  add_predictions(m) |> 
  select(vid, pred) |> 
  distinct()
tidy(m, conf.int = TRUE)
```


#### Considering observers

Looking at the distribution adding the observer we have:

```{r}
d2 |> 
  count(vid, observer, sid) |> 
  ggplot() +
  geom_col(aes(sid, n, fill = observer)) +
  facet_grid(vid ~ ., scales = "free_y") +
  labs(x = "Number of species",
       y = "Number of trips") +
  scale_fill_brewer(palette = "Set1")
```

We observe:

* The number of observer trips is a small proportion of the total trips
* When observer is on board the number of species landed tend to be in the uppper range of the distribution

Could get a quick summary view of the mean and the confidence interval via:

```{r}
d2 |> 
  ggplot(aes(vid, sid, colour = observer)) +
  stat_summary(fun.data = "mean_cl_boot") +
  scale_colour_brewer(palette = "Set1") +
  labs(x = "Vessel",
       y = "Number of species") +
  scale_y_continuous(breaks = 8:20)
```

So, a quick analysis indicates that we have significantly higher number of species landed when observer is on-board. Let's estimate the parameters:

```{r}
m <- glm(formula = sid  ~ vid + observer, data = d2, family = poisson)
summary(m)
```

So again the statistics are significant, both the vessel factor and the observer factor. Nake note that the intercept is now different, was 2.14612 (8.55 species) is now 2.44682 (11.55 species).

The meaning of 11.55 can be considered as the expected mean number of species landed of the first vessel when the observer is on-board, taken into account the data for the second vessel.



#### Adding quarter

```{r}
d2 <- 
  d |> 
  mutate(quarter = quarter(date)) |> 
  filter(vid %in% c(1937, 2170)) |> 
  select(vid, sid = n_sid, q = quarter, observer = Eftirlitsmaður) |> 
  mutate(vid = as.factor(vid),
         observer = as.factor(observer))
```


```{r}
d.count <- 
  d |> 
  group_by(vid) |> 
  summarise(n = n(),
            n.observer = sum(Eftirlitsmaður == "Já"),
            m = median(n_sid)) |> 
  arrange(-m) |> 
  mutate(n = paste0(n, " / ", n.observer))
d |> 
  mutate(vid = factor(vid)) |> 
  ggplot(aes(reorder(vid, n_sid), n_sid, colour = Eftirlitsmaður)) +
  stat_summary(fun.data = "mean_cl_boot") +
  labs(x = "Skipaskárnúmer", y = "Fjöldi landaðra tegunda",
       caption = "Fiskibotnvarpa") +
  scale_colour_brewer(palette = "Set1") +
  coord_flip()
```

Let take some extreme vessels when it comes to median number of species and run a glm-model:

```{r}

m
summary(m)
1 - (m$deviance/m$null.deviance) # variance explained
a <- 
  d2 %>% 
  add_predictions(m) %>% 
  mutate(pred = exp(pred))
a |> 
  ggplot(aes(sid, fill = observer)) +
  geom_histogram(binwidth = 1) +
  facet_grid(vid ~ ., scales = "free_y") +
  scale_fill_brewer(palette = "Set1")

m2 <- glm(formula = sid  ~ vid + observer, data = d2, family = poisson)
m2
summary(m2)
1 - (m2$deviance/m2$null.deviance) # variance explained
```



```{r}
d <- 
  ln |> 
  mutate(year = year(datel),
         month = month(datel)) |> 
  filter(gid_ln == 6) |> 
  select(n_sid, vid, hid = hid_ln, month, observer = Eftirlitsmaður) |> 
  mutate(vid = factor(vid),
         hid = factor(hid),
         month = factor(month),
         observer = factor(observer))
m <- glm(formula = n_sid  ~ vid, data = d, family = poisson)
m <- glm(formula = n_sid  ~ vid + hid + month + observer, data = d, family = poisson)
#m <- glm(formula = n_sid  ~ observer, data = d)
summary(m)
summary(residuals(m))
m$null.deviance
m$deviance
1 - (m$deviance/m$null.deviance)
plot(m)
anova(m, test = "F")

a <- 
  d %>% 
  add_predictions(m) %>% 
  add_residuals(m)
a |> count(pred)
ggplot(a, 
       aes(resid)) + 
  geom_histogram(aes(y=after_stat(density),), position="identity", alpha=0.5, col='grey50', fill='grey50')+
  geom_density(aes(resid),adjust=2.5,color='blue') +
  ylab("Density") + xlab("Model residuals")
mt <- 
  m |> 
  tidy() |> 
  mutate(mean = estimate,
         lower.ci = mean - 2 * std.error,
         upper.ci = mean + 2 * std.error)
mt |> 
  filter(str_starts(term, "hid")) |> 
  ggplot() +
  geom_pointrange(aes(reorder(term, mean), y = mean, ymin = lower.ci, ymax = upper.ci)) +
  coord_flip()
mt |> 
  filter(str_starts(term, "month")) |> 
  mutate(term = str_remove(term, "month"),
         term = as.integer(term)) |> 
  ggplot() +
  geom_pointrange(aes(term, y = mean, ymin = lower.ci, ymax = upper.ci))  +
  scale_x_continuous(breaks = 1:12)
mt |> 
  filter(str_starts(term, "vid")) |> 
  ggplot() +
  geom_pointrange(aes(reorder(term, mean), y = mean, ymin = lower.ci, ymax = upper.ci)) +
  coord_flip()
```

